---
title: "google_bike _rides"
author: "Abo Yehia"
date: "15/09/2021"
output: html_document
---
## loading libraries
```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(janitor)
```

## loading file

```{r}
aug_2020<- read.csv("C:\\Users\\al-khateb\\Desktop\\Cyclic capstone project\\202008-divvy-tripdata.csv")
sep_2020<- read.csv("C:\\Users\\al-khateb\\Desktop\\Cyclic capstone project\\202009-divvy-tripdata.csv")
oct_2020<- read.csv("C:\\Users\\al-khateb\\Desktop\\Cyclic capstone project\\202010-divvy-tripdata.csv")
nov_2020<- read.csv("C:\\Users\\al-khateb\\Desktop\\Cyclic capstone project\\202011-divvy-tripdata.csv")
dec_2020<- read.csv("C:\\Users\\al-khateb\\Desktop\\Cyclic capstone project\\202012-divvy-tripdata.csv")
jan_2021<- read.csv("C:\\Users\\al-khateb\\Desktop\\Cyclic capstone project\\202101-divvy-tripdata.csv")
feb_2021<- read.csv("C:\\Users\\al-khateb\\Desktop\\Cyclic capstone project\\202102-divvy-tripdata.csv")
mar_2021<- read.csv("C:\\Users\\al-khateb\\Desktop\\Cyclic capstone project\\202103-divvy-tripdata.csv")
april_2021<- read.csv("C:\\Users\\al-khateb\\Desktop\\Cyclic capstone project\\202104-divvy-tripdata.csv")
may_2021<- read.csv("C:\\Users\\al-khateb\\Desktop\\Cyclic capstone project\\202105-divvy-tripdata.csv")
june_2021<- read.csv("C:\\Users\\al-khateb\\Desktop\\Cyclic capstone project\\202106-divvy-tripdata.csv")
july_2021<- read.csv("C:\\Users\\al-khateb\\Desktop\\Cyclic capstone project\\202107-divvy-tripdata.csv")
aug_2021<- read.csv("C:\\Users\\al-khateb\\Desktop\\Cyclic capstone project\\202108-divvy-tripdata.csv")
```
## combine files in 1 data frame
```{r}
bike_riders<- rbind(aug_2020, sep_2020, oct_2020, nov_2020,dec_2020,jan_2021,feb_2021,mar_2021,april_2021,may_2021,june_2021,july_2021,aug_2021)
```
## clean empty
```{r}
bike_riders<- janitor::remove_empty(bike_riders,which = c("cols"))
bike_riders<- janitor::remove_empty(bike_riders,which = c("rows"))
```

bike_riders<- janitor::remove_empty(bike_riders,which = c("cols"))
bike_riders<- janitor::remove_empty(bike_riders,which = c("rows"))

## inspect the new data frame
```{r}
View(bike_riders)
```

## convert character type to data/time in start_at , end_at
```{r}
bike_riders$started_at<- lubridate::ymd_hms(bike_riders$started_at)
bike_riders$ended_at<- lubridate::ymd_hms(bike_riders$ended_at)
```

## convert character type to numeric in start_station_id , end_station_id
```{r}
bike_riders$start_station_id<- as.numeric(bike_riders$start_station_id)
bike_riders$end_station_id<- as.numeric(bike_riders$end_station_id)
```

##  In the "member_casual" column, there are two names for members and two names for casual riders
```{r}
bike_riders <- bike_riders %>% 
  mutate(member_casual= recode(member_casual,"Subscriber" = "member", "Customer"= "casual"))
```

## add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate the data.
```{r}
bike_riders$date <- as.Date(bike_riders$started_at)
bike_riders$month<- format(as.Date(bike_riders$date),"%m")
bike_riders$day<- format(as.Date(bike_riders$date),"%d")
bike_riders$year<- format(as.Date(bike_riders$date),"%Y")
bike_riders$day_of_Week<- format(as.Date(bike_riders$date),"%a")
## calculate ride length in seconds
bike_riders$ride_length<- difftime(bike_riders$ended_at,bike_riders$started_at,units = "mins")
```

## Inspect the structure of the columns
```{r}
str(bike_riders)

```

## remove "bad" data
```{r}
bike_riders_v2 <- bike_riders[!(bike_riders$ride_length<0),]
View(bike_riders_v2)
```

## descriptive analysis on ride_length 
```{r}
mean(bike_riders_v2$ride_length)
median(bike_riders_v2$ride_length)
max(bike_riders_v2$ride_length)
min((bike_riders_v2$ride_length))
```

## compare members and casual users
```{r}
aggregate(bike_riders_v2$ride_length ~ bike_riders_v2$member_casual, FUN = mean)
aggregate(bike_riders_v2$ride_length ~ bike_riders_v2$member_casual, FUN = median)
aggregate(bike_riders_v2$ride_length ~ bike_riders_v2$member_casual, FUN = max)
aggregate(bike_riders_v2$ride_length ~ bike_riders_v2$member_casual, FUN = min)
```

## average ride time by each day for members vs casual users
```{r}
aggregate(bike_riders_v2$ride_length ~ bike_riders_v2$member_casual + bike_riders_v2$day_of_Week,FUN = mean)
bike_riders_v2$day_of_Week<- ordered(bike_riders_v2$day_of_Week,levels=c("Sunday","Monday","Tuesday","Wednesday","Thursday","Friday","Saturday"))
```

## analyze ridership data by type and weekday
```{r}
bike_riders_v2 %>% 
  mutate(weekday =wday(started_at , label = TRUE)) %>% 
  group_by(member_casual,weekday) %>% 
  summarise(number_of_rides = n(),average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)
```

## visualize the number of rides by rider 
```{r}
library(scales)
bike_riders_v2 %>% 
  mutate(weekday =wday(started_at , label = TRUE)) %>% 
  group_by(member_casual,weekday) %>% 
  summarise(number_of_rides = n(),average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge"  ) +labs(title =  "Number of rides by rider type", caption = "figure 1")+ 
  scale_y_continuous(labels = comma)
```
### The figure 1 shows that the bike usage of annual members are relatively stable throughout the week, while casual riders tend to spike up during Saturdays and Sundays

## create a visualization for average duration
```{r}
bike_riders_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") + labs(title =  "Average Duration by Weekday" , caption = "figure 2")

```
### The figure 2 shows that Throughout the week, casual riders tend to use the bikes a lot more compared to annual


## Key Takeaways:

* Annual members consistently use the bikes during the whole week, while casual riders tend to spike on weekends (Saturday and Sunday)
* Casual riders tend to use the bikes for a lot longer duration compared to annual members in all days of the week.

## Recommendations:
* Create an incentive program for annual members that rewards ride duration over the number of rides.
* Add incentives for annual members who use the bikes in the weekends.
